---
title: "Assignment for Week 6"
author: "Author: Your name here!"
date: "2/3/2017"
output: pdf_document
---
<!-- Please turn in a PDF version of this assignment with answers.  For help, see the template.--> 

```{r setup, include=FALSE}
# This code changes "global" settings, here it makes the computer `echo` which refers to 
# printing the code as well as the output into the pdf document when you knit.

knitr::opts_chunk$set(echo = TRUE)

# Change the following code to be your working folder for R

knitr::opts_knit$set(root.dir = '~/Dropbox/mathCamp/Code/data/')
# The following allows the computer to always get the same result for random stuff.
set.seed(100)

```

# Assignment
<!-- Please Put \newpage commands between long answers.  Avoid allowing answers to spill over pages. --> 
1)  2.1 (p. 119)

2)  2.3 (p. 119)

3)  2.16 (p. 122)

4)  2.20 (p. 123)

5)  3.1 (p. 184)

6)  3.2 (p. 184) (Use integration by parts)

7)  3.8 (b) (p. 186)

8)  3.11 (a) (b) (c) (p. 188)


9) Plot the pmf of each distribution in the following list, with at least one variation in the listed parameter(s), using either a histogram or with a custom `table` and a `persp` plot.

- bernoulli, $p$
- binomial, $p$
- Poisson, $\lambda$
- Geometric, $p$
- Hypergeometric, $m$, $n$
- negative binomial, $p$


10) Using `str()` and the R help file and worksheet, describe in words each of the elements of the object chisqt1.
```{r}
x <- c(89,37,30,28,2)
p <- c(.40,.20,.20,.15,.05)
chisqt1 <- chisq.test(x, p = p)
```



