---
title: "Assignment for Week 3"
author: "Author: Your name here!"
date: "12/14/2016"
output: pdf_document
---
<!-- Please turn in a PDF version of this assignment with answers.  For help, see the template.--> 

```{r setup, include=FALSE}
# This code changes "global" settings, here it makes the computer `echo` which refers to 
# printing the code as well as the output into the pdf document when you knit.

knitr::opts_chunk$set(echo = TRUE)

# Change the following code to be your working folder for R

setwd("~/Dropbox/mathCamp/Code/R")
# The following allows the computer to always get the same result for random stuff.
set.seed(100)

# install.packages("dplyr")
# install.packages("ggplot2")
# install.packages("nycflights13")
```

## Package Loading

```{r Package Loading}
library(dplyr, warn.conflicts = FALSE)
library(ggplot2)
library(haven)
```

# Assignment
<!-- Please Put \newpage commands between long answers.  Avoid allowing answers to spill over pages. --> 

1) 19.2.3 (a) (p. 405)
2) 19.2.6 (p. 406)
3) 11.1.4 (c, e, f) (p. 207)
4) 11.2.4 (p. 213)
5) 11.2.6 (p. 213)
6) 11.3.1 (p. 217)

\newpage


7) Which of the following plots of the same data is clearer?  Why? 

```{r, echo=FALSE, fig.width=3.5, fig.height=3}
library(ggplot2)
df<- data.frame(y=c(10,11,13,12,14), Answers=c("Very bad", "Somewhat bad","Not great", "Mediocre","Average" ))
ggplot(df, aes(x=factor(1), y=y, fill=Answers))+geom_col()+coord_polar(theta="y")+xlab("")+ylab("")+theme_bw()+theme(axis.line=element_blank(),axis.text.x=element_blank(),
          axis.text.y=element_blank(),axis.ticks=element_blank(),
          axis.title.x=element_blank(),
          axis.title.y=element_blank(),
          panel.background=element_blank(),panel.border=element_blank(),panel.grid.major=element_blank(),
          panel.grid.minor=element_blank(),plot.background=element_blank())+ggtitle("Plot A")

ggplot(df, aes(x=factor(1), y=y, fill=Answers))+geom_col(position="dodge")+xlab("")+ylab("")+theme_bw()+theme(axis.line=element_blank(),axis.text.x=element_blank(),
          axis.text.y=element_blank(),axis.ticks=element_blank(),
          axis.title.x=element_blank(),
          axis.title.y=element_blank(),
          panel.background=element_blank(),panel.border=element_blank(),panel.grid.major=element_blank(),
          panel.grid.minor=element_blank(),plot.background=element_blank())+ggtitle("Plot B")


```

\vspace{1 cm}
8)  Use the Afghanistan dataset to answer the following:

    - What if we take the average elevation of each language, which is spoken in the second highest altitudes on average?
    - What language's speakers are farthest away from Kabul on average?
    - (Extra Credit) What is the most geographically isolated city with a population more than 1000 in Afghanistan?

9) Correct the common mistakes in the following code snippets

  a)  Make the following TRUE
  
      ```
      mean(c(3, 2, 1, NA)) == 2
      ```
  b)  Make the following TRUE
  
      ```
      df <- data.frame(var1 = c(35, 40, 45));
      length(df[1])==3
      ```
  c)  Make the following TRUE
  
      ```
      df <- data.frame(var1 = c(35, 40, 45));
      dplyr::filter(df, "var1"==40)==40
      ```
  d)  Make the following produce a plot of the points described in x and y.
  
      ```
      x <- c(3, 10, 4);
      y <- c(2, 3, 4);
      ggplot()+geom_point(x, y)
      ```
  e)  Produce the string `Iowa' 
    ```
    library(dplyr);
    df <- data.frame(var1 = c(35, 40, 45), var2=c("Iowa", "France","Mexico"));
    dfs <- df%>%filter(var2="Iowa")%>%select("var2");
    as.character(dfs[[1]])
    ```
    
10)  Make a plot of Afghanistan which tells us about the geography of language diversity.  Avoid overplotting, where the points overlap to much.  Make sure to label your x and y axis, and to make a nice title.  Check out https://cran.r-project.org/web/packages/ggthemes/vignettes/ggthemes.html and http://shinyapps.org/apps/RGraphCompendium/index.php for ideas.


```{r, include=FALSE, eval=FALSE}
# SOLUTION
NlanguagesPerRegion <-
  afg %>% group_by(MISTI_PROV_CODE) %>% summarise(ndistinct = n_distinct(LANGUAGE))
  
  afg <- afg %>% left_join(NlanguagesPerRegion)
  
  pdf("~/Google Drive/ChicagoSyllabi/SSI131/AfghanistanOut.pdf",
  width = 10)
  ggplot(afg, aes(LON_X, LAT_Y, color = ndistinct)) + geom_point(size = .1) +
  guides(color = guide_legend(override.aes = list(size = 5))) +
  theme_bw() +
  ggtitle("Provincial Language Diversity in Afghanistan")
  dev.off()

```