---
title: "Assignment for Week 6"
author: "Author: Your name here!"
date: "2/3/2017"
output: pdf_document
---
<!-- Please turn in a PDF version of this assignment with answers.  For help, see the template.--> 

```{r setup, include=FALSE}
# This code changes "global" settings, here it makes the computer `echo` which refers to 
# printing the code as well as the output into the pdf document when you knit.

knitr::opts_chunk$set(echo = TRUE)

# Change the following code to be your working folder for R

knitr::opts_knit$set(root.dir = '~/Dropbox/mathCamp/Code/data/')
# The following allows the computer to always get the same result for random stuff.
set.seed(100)

```

## Package Loading

```{r Package Loading}
library(ggplot2)

```

# Assignment
<!-- Please Put \newpage commands between long answers.  Avoid allowing answers to spill over pages. --> 
1) Let $E, F,$ and $G$ be events.  Find the expressions (using $\cap$ $\cup$ and $^c$) for the events that of $E, F,$ and $G$
  a) only $F$ occurs,
  b) both $E$ and $F$ but not $G$ occur,
  c) at least one event occurs,
  d) at least two events occur,
  e) all three events occur,
  f) none occurs,
  g) at most one occurs,
  h) at most two occur.
  
2)  1.17 (p. 57)

3)  1.19 (p. 57)

4)  1.20 (a) (b) (p. 57) 

5)  1.39 (p. 62)

6)  1.56 (a) (b) (p. 68)

\newpage

7) Use R to simulate the following problems:

(a) If three players are dealt five cards from a standard 52 card deck, how what is the probability that at least one person gets exactly two aces?

(b) Suppose that $n$ letters are placed randomly in $n$ envelopes, what is the probability that $n-1$ letters will be placed in the correct envelope?

(c) Suppose that $n$ letters are placed at random in $n$ envelopes, if $q_n$ is the probability that no letter is placed in the correct envelope, which of the following has the largest $q_n$? $n=10$, $n=21$, $n=53$, $n=100$.  

(d) If the $n$ letters are placed in $n$ envelopes, what is the expected number of letters that go to the correct destination if $n=10$, $n=20$, $n=50$?  

8) Plot the pmf of each distribution in the following list, with at least one variation in the listed parameter(s), using either a histogram or with a custom `table` and a `persp` plot.

- bernoulli, $p$
- binomial, $p$
- Poisson, $\lambda$
- Geometric, $p$
- Hypergeometric, $m$, $n$
- negative binomial, $p$
- Multinomial, size, $p=(p_1,p_2)$

9) Using `str()` and the R help file and worksheet, describe in words each of the elements of the object chisqt1.
```{r}
x <- c(89,37,30,28,2)
p <- c(.40,.20,.20,.15,.05)
chisqt1 <- chisq.test(x, p = p)
```



