---
title: "Assignment for Week 4"
author: "Author: Your name here!"
date: "1/22/2016"
output: pdf_document
---
<!-- Please turn in a PDF version of this assignment with answers.  For help, see the template.--> 

```{r setup, include=FALSE}
# This code changes "global" settings, here it makes the computer `echo` which refers to 
# printing the code as well as the output into the pdf document when you knit.

knitr::opts_chunk$set(echo = TRUE)

# Change the following code to be your working folder for R

setwd("~/Dropbox/mathCamp/Code/R")
# The following allows the computer to always get the same result for random stuff.
set.seed(100)

# install.packages("dplyr")
# install.packages("ggplot2")
# install.packages("nycflights13")
```

## Package Loading

```{r Package Loading}
library(dplyr, warn.conflicts = FALSE)
library(ggplot2)
library(haven)
```

# Assignment
<!-- Please Put \newpage commands between long answers.  Avoid allowing answers to spill over pages. --> 

1. 19.2.3 (a) (p. 405)

2. 19.2.6 (p. 406)

3. 11.1.4 (c, e, f) (p. 207)

4. 11.2.4 (p. 213)

5. 11.2.6 (p. 213)

6. 11.3.1 (p. 217)

\newpage

7) Which of the following datasets are tidy?

```
dta1<- data.frame(year = c(2000:2003,2000:2003), 
                  CountryA = c(rep("USA",4),rep("CAN",4) ),
                  CountryB=c(rep("Russia",4),rep("Russia",4) ),
                  Agreement=c(0,1,1,0,0,1,0,0))

dta2<- data.frame(year = c(2000:2003,2000:2003), 
                  Dyad = c(rep("USA-RUS",4),rep("USA-RUS",4) ),
                  ConflictType=c(rep("War",4),rep("Minor MID",4) ),
                  ConflictLevel=c(0,3,1,0,0,1,0,0))

dta3<- data.frame(Dyad = c("USA-RUS", "CAN-RUS"),
                  War2002=c(0,2),
                  War2003=c(0,2),
                  War2004=c(0,2),
                  MinorMID2002=c(0,2),
                  MinorMID2003=c(0,2),
                  MinorMID2004=c(0,2))
```
8) Use spread, gather and seperate from the tidy package to fix the untidy data.frames from question 7.

9) Correct the common mistakes in the following code snippets

  a)  Make the following TRUE
  
```{}
mean(c("3,000", 2, 1, 0, 4)) == 601.4
```
  b)  Make the following into a vector TRUE TRUE TRUE
  
```{}
EuclidDistancetoKabul <- function(x) {
  sqrt((-64 - x[1]) ^ 2 + (16 - x[2]) ^ 2)
}
x <- EuclidDistancetoKabul(rbind(c(48, 4), c(12, 3), c(80, 13)))
x > 20
```
      
      
  c)  Make the following code install the 'car' package.
  
```{}
install.packages(car)
```
      
  d)  Correct the following plot command:
  
```{}
x <- c(3, 10, 4)
y <- c(2, 3, 4)
df <- data.frame(x, y)
ggplot(df) + geom_bar(aes(x, y))
```
      
  e)  Make the following code report TRUE 
  
```{}
allthetypes <- list(df1 = data.frame(x = c("Iowa"), stringsAsFactors = F))
is.character(allthetypes[1][[1]])
```
\newpage

10)  How many shootings occur in the state with the most active VA hospitals?  The least active?

11)  Plot the following function for $x \in [-4,4]$ by writing a custom function().  Use integrate() on that function to plot the area under the curve of that function from -Inf to x, such that $x \in [-4,4]$.

$$\frac{1}{\sqrt{2\pi}}e^{\frac{-x^2}{2}}$$






