---
title: "Assignment for Week 1"
author: "Author: Your name here!"
date: "12/14/2016"
output: pdf_document
---
<!-- Please turn in a PDF version of this assignment with answers.  For help, see the template.--> 

```{r setup, include=FALSE}
# This code changes "global" settings, here it makes the computer `echo` which refers to 
# printing the code as well as the output into the pdf document when you knit.

knitr::opts_chunk$set(echo = TRUE)

# Change the following code to be your working folder for R

setwd("~/Dropbox/mathCamp/Code/R")
# The following allows the computer to always get the same result for random stuff.
set.seed(100)

# install.packages("dplyr")
# install.packages("ggplot2")
# install.packages("nycflights13")
```

# Assignment Part I
<!-- Please Put \newpage commands between long answers.  Avoid allowing answers to spill over pages. --> 
\begin{itemize}

\item[3.13] Let $A, B, C$ be subsets of a set $X$.  Which of the expressions $(A\cap B)\cup (A\cap B^c)$, $(A\cap B)\cup (A\cap B^c)^c$, $(A\cap B)\cup (A^c\cup C)^c$ can be simplified, and how?



\item[3.4.2] Let $g$ be the mapping from $\mathbb{R}^2$ to $\mathbb{R}^2$ defined by $$g(x, y)=(y, -x),$$ and let the mapping $h$ be as in Exercise 3.4.1.  Find the image of $(x, y)$ under each of the composite mappings $g\circ h$ and $h\circ g$.


\item[31.1.1] As in the text, let $\mathbb{Q}$ be the set of all rational numbers. If $x\in \mathbb{Q}$, which of the following statements (P, Q, R, S) imply which of the others?

\begin{itemize}

\item[P:] $x\geq 0$
\item[Q:] $\exists w\in \mathbb{Q} $ such that $x=w^2$
\item[R:] $nx>-1$ $\forall n\in \mathbb{N}$
\item[S:] $\exists n\in \mathbb{N}$ such that $nx>-1$.

\end{itemize}

\item[5.1.3] In each of the following cases, state whether the sequence $\{u_n\}$ tends to a limit, and find the limit if it exists:

\begin{itemize}
\item[(a)] $u_n=1+\frac{1}{2}n$
\item[(b)] $u_n=1-\frac{1}{2}n$
\item[(c)] $u_n=(\frac{1}{2})^n$
\item[(d)] $u_n=-(\frac{1}{2})^n$
\end{itemize}

\end{itemize}



\pagebreak

# Assignment Part II

## Package Loading

```{r Package Loading}
library(dplyr, warn.conflicts = FALSE)
library(ggplot2)
library(nycflights13)
```

(1) What are the types of data that serve as the argument and value of the following functions?

* `nchar()`
* `log()`
* `paste()`
* `is.na()`

(2) What was the longest physical distance taken out of New York in 2013? 

(3) In R, we can use scientific notation.  2e3 is 2000.  Why is log(1e10)-log(1e8) the same as log(1e2)?

(4) Without running these commands, what are the values of C, D, and E?

```{r LogicQuestion, eval=FALSE }
A <- "Iowa"
B <- "Kansas"

C <- !((A == "Kentucky") | (B == "Kansas"))

D <- ((nchar(A) > 2) & is.na(B)) | (!is.na(A))

E <- ifelse( C, D, A)
```


(5) Plot an interesting subset of the data Flights data.  Use the Rmarkdown fig.height and fig.width to make it look nice.

(6) Write your own "quantile" function which gets the median and the 95th percentile of data.  

(7) What does `df` look like when it is printed out?

```{r Indexing Question, eval=FALSE }
df <- data.frame(dogheight=c("10", "20", "30"), dogwidth=c("5", "3", "1"))
df[df$dogwidth==3,1] <- NA
```

(8) _True_ or _False_

    (a) A small p-value, such as .001 or .0001, means that our estimate is unlikely to have arisen from the null hypothesis.
    (b) The small errors distribution is a nice approximation for the distances traveled by flights out of New York City.
    (c) $f(x) = 2x$ for $0\leq x \leq 1$ is a possible probability density function. 
    (d) The probability that a normal distribution produces the number 0 is 0.
