---
title: "Assignment for Week 8"
author: "Author: Your name here!"
date: "2/20/2017"
output: pdf_document
---
<!-- Please turn in a PDF version of this assignment with answers.  For help, see the template.--> 

```{r setup, include=FALSE}
# This code changes "global" settings, here it makes the computer `echo` which refers to 
# printing the code as well as the output into the pdf document when you knit.

knitr::opts_chunk$set(echo = TRUE)

# Change the following code to be your working folder for R

knitr::opts_knit$set(root.dir = '~/Dropbox/mathCamp/Code/data/')
# The following allows the computer to always get the same result for random stuff.
set.seed(100)

```

## Package Loading

```{r Package Loading}
library(ggplot2)
library(dplyr, warn.conflicts = F)
```

# Assignment
<!-- Please Put \newpage commands between long answers.  Avoid allowing answers to spill over pages. --> 
1)  5.1 (p 284)

2)  5.8 (p 290)

3)  9.20 (p 518)

4) 9.22 (a) (b) (p 519)

5) 9.26 (p 519)

6) 9.27 (p 519)

7) 9.30 (p 520)

8) Suppose $\bar{X}$ is the mean of 100 observations from a popualtion
  with mean $\mu$ and variance $\sigma^2$.  Find limits between which
  $\bar X - \mu$ will lie with a probability of at least .85 using the Central Limit Theorem.
  
9) Consider the function $f(x,y)$ defined by the following table

|  |y= 1|  2  |3   |
|------:|:------:|:------:|:------:|
|x= 1  | .1  |    0   |   .1  |
|  2  |  0  |   .2|   .2 |
|   3|    .1  |     .2   | .1  |

Table: $f(x,y)$

(a) Verify that $f(x,y)$ is a PDF.
(b) Find the marginal probability functions of X and Y.
(c) Compute E(X) and V(X).
(d) Compute the covariance between X and Y. 
(e) Find $P(X<Y)$.
(f) Are X and Y independent? Justify your answer.


\newpage

# R assignment

10) Write your own functional equivalents to the listed values of each of the following functions.  Each should take on a vector and output one of the listed values.

a) `t.test(X,Y, var.equal=TRUE)` - statistic, parameter, p.value, conf.int.

b) `var.test(X,Y)` - statistic, parameter, p.value, conf.int.

