---
title: "Assignment for Week 7"
author: "Author: Your name here!"
date: "2/20/2017"
output: pdf_document
---
<!-- Please turn in a PDF version of this assignment with answers.  For help, see the template.--> 

```{r setup, include=FALSE}
# This code changes "global" settings, here it makes the computer `echo` which refers to 
# printing the code as well as the output into the pdf document when you knit.

knitr::opts_chunk$set(echo = TRUE)

# Change the following code to be your working folder for R

knitr::opts_knit$set(root.dir = '~/Dropbox/mathCamp/Code/data/')
# The following allows the computer to always get the same result for random stuff.
set.seed(100)

```

## Package Loading

```{r Package Loading}
library(ggplot2)
library(dplyr, warn.conflicts = F)
```

# Assignment
<!-- Please Put \newpage commands between long answers.  Avoid allowing answers to spill over pages. --> 
1)  4.17 (p 249)

2)  4.18 (p 249)

3)  4.19 (p 249)

4)  4.30 (p 256)

5)  4.31 (p 256)


\newpage

R assignment

6)  Plot the number of bombs dropped in each contemporary European country during World War I.

7)  Estimate the location (coordinates) of the center of gravity of bombs dropped during World War I using the Thor dataset.  Each row of the Thor dataset describes many bombs being dropped.  How does that average move over the course of the war?  Estimate its location for 1, 3, and 6 month rolling averages.  




