---
title: "File Paths in R"
author: "Robert Gulotty"
date: "September 30, 2014"
output: pdf_document
---

## File Paths
This is a tutorial on file paths and directories (folders).

A directory indicates the location on the computer that you are saving your information, usually on your hard drive.  A path is a slash-separated list of directory names, followed by a file name or a directory (folder) name.  For example, `/Users/robertgulotty` suggests that there is a folder named `robertgulotty` in the `Users` folder.  

If you look at the ___file path___ of some file in your computer, by right clicking on the file and selecting properties or get info, you will find that directory names are separated by `/` in Macs and other Unix based systems and `\` in Windows (PC).

These slashes differ across Macs and PCs, and R has trouble with backslashes (`\`).  If you use R on a mac or a PC, you will see the path printed with a forwardslash (`/`).  This is because R is built for Unix, and backslash is a programming command in the R language.  

On Windows machines to tell your computer where to look, you must either 

+ use the forwardslash (/)
+ use ___two___ backslashes (\\\\).

Either one will work.

## Working Directory

The following command tells us what the `working directory' is:
```{r}
getwd()
```

A working directory is the default place for R to look for files and folders.  If you read out data into a csv, you will create a file in the working directory.  

If my working directory is "/Users/robertgulotty/Google Drive/stuffforclass/" the following three commands are the same:
```{r}
# WBdta<-read.csv("/Users/robertgulotty/Google Drive/stuffforclass/WB_dta.csv")
# WBdta<-read.csv("WB_dta.csv")
# WBdta<-read.csv("~/Google Drive/stuffforclass/WB_dta.csv")
```


In the Mac operating systems, a directory usually begins with `/Users/yourname`  This can be replaced with a $\sim$, which is a shortcut for `/Users/yourname`

In Windows operating systems, a directory usually begins with 
`C:/Users/yourname/`

\newpage

## setwd() to change the Working Directory

You can change directories as follows:

```{r}
## Mac change directory
# setwd("/Users/robertgulotty/Google Drive/IPS2014/")
## PC change directory (Version 1)
# setwd("C:/Users/Google Drive/IPS2014/")
## PC change directory (Version 2)
# setwd("C:\\Users\\Google Drive\\IPS2014\\")

```
Where the bit in quotation markes is replaced by your preferred file.  Note that you can use the tab key to complete the file name!

----------------------------------  
__Expert Tip__

If you need to share code, and you are often switching between different operating systems, remember to set your working directory at the top of the R script.  This will avoid having to write the path later on.
----------------------------------