---
title: "Regression Presentation"
author: "Robert Gulotty"
date: "11/6/2020"
output: beamer_presentation
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
library(car)
data(Prestige)
```

## The Ordinary Least Squares Model

In many quantitative papers people display standard outputs from statistical software.  Today we will cover three main ideas.

- How are are the parameters of the model ($\beta$) calculated.
- How are standard errors calculated and summarized
- What diagnostics are reported.

## Prestige Data

```{r Prestige, echo = TRUE}
summary(Prestige[,c("prestige", "education", "income")])
```


## Predicting Prestige with lm()
\footnotesize
```{r reg with LM, echo=TRUE}
summary(lm(prestige~education+income, data=Prestige))
```

## Visulazation of Prestige
```{r pic}
scatter3d(prestige~income+education, data=Prestige)
```

## Output and Input
```{r outputs, echo=TRUE}
y<-Prestige$prestige
xvars<-as.matrix(Prestige[,c("education", "income")])
# Add vector of 1 to X matrix
xmat<-cbind(1,xvars)
```



## $\beta=(X'X)^{-1}X'y$.
```{r point estimates, echo=TRUE}
betas<-solve(t(xmat)%*%xmat)%*%t(xmat)%*%y
betas
```

## Note Betas
\footnotesize
```{r betas in lm, echo=TRUE}
summary(lm(prestige~education+income, data=Prestige))
```



## Errors= $y-X\beta$.
```{r errors, echo=TRUE}
err<-y-xmat%*%betas
summary(err)
```
## Note errors
\footnotesize
```{r errors in lm, echo=TRUE}
summary(lm(prestige~education+income, data=Prestige))
```

## Residual standard error: sqrt(e'e/(n-k))
\footnotesize
```{r rse, echo=TRUE}
ete<-t(err)%*%err/(length(err)-ncol(xmat))
sqrt(ete)
```

## Note Residual standard error
\footnotesize
```{r resid errors in lm, echo=TRUE}
summary(lm(prestige~education+income, data=Prestige))
```

## $R^2$: 1-sum($(y-xb)^2$)/sum($(y-ybar)^2$)
```{r rsquared, echo=TRUE}
1-sum(err^2)/sum((y-mean(y))^2)
```

## Variance Covariance Matrix Betas
```{r varcovar, echo=TRUE}
(varcovar<-as.numeric(ete)*solve(t(xmat)%*%xmat))
```

## Standard Errors of Betas
```{r beta se, echo=TRUE}
sigmas<-diag(varcovar)
(stderrors<-sqrt(sigmas))
```

## Note Standard Errors
\footnotesize
```{r sd error, echo=TRUE}
summary(lm(prestige~education+income, data=Prestige))
```

## tstats of Betas

```{r tstats, echo=TRUE}
(tstats<-betas/stderrors)
```

## P-values of Betas

```{r pvals, echo=TRUE}
(1-pt( abs(tstats), df= length(err)-3))*2
```


